---
title: "Corn Earworm (CEW) Analysis"
author: "Evan Dadson and Andrew Keziah"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
mainfont: Times New Roman
fontsize: 12pt
geometry: margin=1in
---

# Introduction to R

R can be used in the command line, in scrips with the .R extension, or in integrated development environments (IDEs) (e.g. RStudio, Jupyter Notebooks, Jupyterlab, etc.). The easiest way to use R is with RStudio. An IDE allows you to write code and test output all in one place, and is more robust than other methods (e.g. shell scripting in the terminal).

####R as a Calculator

```{r}
1+1
```

```{r}
x=3
x*5
```

```{r}
# This is a comment
# Comments aren't processed like lines of code
# This provides clarity for people to review and repeat your code
```

```{r}
# In most programming languages, you must end an executable line of code with a
# semicolon. 
# In R this is optional.
x = 6;
x = 5; y = 7 # you can execute multiple statements on the same line
```

Practice by uncommenting this line of code. Does it work?
```{r}
# y = 5 y = 7
```

## Variables and operations

There are a few data types in R. The main ones we care about are integers, characters, and doubles.

```{r}
a <- 12 # this is an integer
b <- "Hello, world!" # this is a character
c <- 12.12 # this is a double
```

Now that we have created some variables using our basic data types, let's print them out to the console.
```{r}
a
b
c
# or we can use better printing
print(b)
print(paste(b, "In the U.S., you finish high school in the", a, "th grade."))
```

Let's jump into some more complex data structures: vectors, matricies, and lists. First, vectors are list of elements that are of the same length. For example, (1, 2, 3, 4), is a vector of integers and is length 4. In R, we can make the previous vector like so,
```{r}
c(1,2,3,4)
# now we can give this vector a name
first_vector <- c(1,2,3,4)
class(first_vector) # what is the data type?
str(first_vector) # what is the data structure?
length(first_vector)
# we can see that we have created a numeric vector of length 4.
```

What if we wanted to make a matrix. This can be tought of as a vector that is of size n x m.

## Analysis
To start our analysis we first need to examine the data. 
```{r}
data_dir <- "/Volumes/cmjone25/Data/Original/pest-occurrence/corn_earworm/"
output_dir <- "/Volumes/cmjone25/Data/Raster/USA/pops_casestudies/corn_earworm/outputs"
trap_data = read.csv(paste0(output_dir, "/traps_extracted_data.csv"))
```

```{r}
head(trap_data)
dim(trap_data)
```

```{r}
lm_fit <- lm(cum_CEW_count ~ percent_corn + percent_cotton + percent_peanuts + percent_soybeans + percent_sorghum + percent_sweetcorn + percent_tobacco + egg_area + factor(year), data = trap_data)
summary(lm_fit)
```

## Visualization

### base R plotting

The R programming language has a number of built in functions we can use to plot and visualize things.
```{r}
trap_data_2020 <- trap_data[trap_data$year == 2020, ]
trap_data_2021 <- trap_data[trap_data$year == 2021, ]
trap_data_2022 <- trap_data[trap_data$year == 2022, ]
trap_data_2023 <- trap_data[trap_data$year == 2023, ]
rbPal <- colorRampPalette(c('blue', 'red'))
intervals_2020 <- cut(trap_data_2020$cum_CEW_count, breaks = 5)
col_vals_2020 <- rbPal(10)[as.numeric(intervals_2020)]
intervals_2021 <- cut(trap_data_2021$cum_CEW_count, breaks = 5)
col_vals_2021 <- rbPal(10)[as.numeric(intervals_2021)]
intervals_2022 <- cut(trap_data_2022$cum_CEW_count, breaks = 5)
col_vals_2022 <- rbPal(10)[as.numeric(intervals_2022)]
intervals_2023 <- cut(trap_data_2023$cum_CEW_count, breaks = 5)
col_vals_2023 <- rbPal(10)[as.numeric(intervals_2023)]
```

```{r, fig.asp=1}
# 2020
plot(trap_data_2020$long, trap_data_2020$lat, col = col_vals_2020, pch = 20,
     cex = 3, xlab = "Longitude", ylab = "Latitude", 
     main = "Cumulative CEW Catch 2020")
legend("bottomright", title="catch",legend=levels(intervals_2020),
       col=rbPal(10), pch=20)
```

```{r, fig.asp=1}
# 2021
plot(trap_data_2021$long, trap_data_2021$lat, col = col_vals_2021, pch = 20,
     cex = 3, xlab = "Longitude", ylab = "Latitude", 
     main = "Cumulative CEW Catch 2020")
legend("bottomright", title="catch",legend=levels(intervals_2021),
       col=rbPal(10), pch=20)
```

```{r, fig.asp=1}
# 2022
plot(trap_data_2022$long, trap_data_2022$lat, col = col_vals_2022, pch = 20,
     cex = 3, xlab = "Longitude", ylab = "Latitude", 
     main = "Cumulative CEW Catch 2020")
legend("bottomright", title="catch",legend=levels(intervals_2022),
       col=rbPal(10), pch=20)
```

```{r, fig.asp=1}
# 2023
plot(trap_data_2023$long, trap_data_2023$lat, col = col_vals_2023, pch = 20,
     cex = 3, xlab = "Longitude", ylab = "Latitude", 
     main = "Cumulative CEW Catch 2020")
legend("bottomright", title="catch",legend=levels(intervals_2023),
       col=rbPal(10), pch=20)
```

Wow! That was a lot of writing to get these plots. They are good, but they could be a whole lot better. The `ggplot2` package has a lot of helpful tools for better and more visually appealing plots.

### ggplot2 plotting

#Backup
```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = cum_CEW_count), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = percent_corn), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = percent_soybeans), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = percent_cotton), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = percent_peanuts), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = percent_tobacco), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = ), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

```{r, fig.asp = 1}
library(ggplot2)
nc_map=map_data("county", region = "north carolina")
ggplot(trap_data, aes(x=long, y=lat)) + 
  geom_polygon(data = nc_map, aes(x=long, y=lat, group=group), 
               linewidth = 1, col = "grey40", fill = NA) + 
  geom_point(aes(colour = cum_CEW_count, size = egg_area), shape = 20) + 
  scale_colour_gradientn(colors = rbPal(10), name = "cumulative CEW \nCatch",
                        na.value = NA) + 
  coord_quickmap(xlim = c(min(trap_data$long) - 0.1, max(trap_data$long) + 0.1),
                 ylim = c(min(trap_data$lat) - 0.1, max(trap_data$lat) + 0.1)) + 
  facet_wrap(~year)
```

#New Buffer Hard-Code - 200m
```{r}
for (year in seq(2020, 2023)) {
  cdl_data <- rast(paste0(cdl_path, year, "_30m_cdls/", year, "_30m_cdls.tif"))
  cdl_data <- crop(cdl_data, empty_vect)
  trap_year <- trap_vect[trap_vect$year == year,]
  trap_buffer <- buffer(trap_year, 200)
  trap_extract <- terra::extract(cdl_data, trap_buffer)
  trap_year$percent_corn <- table(trap_extract)[,c("Corn")]/nrow(trap_extract)
  trap_year$percent_cotton <- table(trap_extract)[,c("Cotton")]/nrow(trap_extract)
  trap_year$percent_peanuts <- table(trap_extract)[,c("Peanuts")]/nrow(trap_extract)
  trap_year$percent_soybeans <- table(trap_extract)[,c("Soybeans")]/nrow(trap_extract)
  trap_year$percent_sorghum <- table(trap_extract)[,c("Sorghum")]/nrow(trap_extract)
  trap_year$percent_sweetcorn <- table(trap_extract)[,c("Sweet Corn")]/nrow(trap_extract)
  trap_year$percent_tobacco <- table(trap_extract)[,c("Tobacco")]/nrow(trap_extract)
  trap_year$egg_area <- sum(table(trap_extract)[,c("Corn")],
                            table(trap_extract)[,c("Cotton")],
                            table(trap_extract)[,c("Peanuts")],
                            table(trap_extract)[,c("Soybeans")],
                            table(trap_extract)[,c("Sorghum")],
                            table(trap_extract)[,c("Sweet Corn")],
                            table(trap_extract)[,c("Tobacco")])
  assign(paste0("trap_", year), trap_year)
}

trap_vect <- rbind(trap_2020, trap_2021, trap_2022, trap_2023)
trap_vect <- project(trap_vect, "EPSG:4326")
trap_data <- cbind(as.data.frame(trap_vect), long = crds(trap_vect)[,1],
                   lat = crds(trap_vect)[,2])
write_csv(trap_data, file = paste0(output_dir, "/traps_extracted_data.csv"))
```

# Buffer 500
```{r}
for (year in seq(2020, 2023)) {
  cdl_data <- rast(paste0(cdl_path, year, "_30m_cdls/", year, "_30m_cdls.tif"))
  cdl_data <- crop(cdl_data, empty_vect)
  trap_year <- trap_vect[trap_vect$year == year,]
  trap_buffer <- buffer(trap_year, 500)
  trap_extract <- terra::extract(cdl_data, trap_buffer)
  trap_year$percent_corn <- table(trap_extract)[,c("Corn")]/nrow(trap_extract)
  trap_year$percent_cotton <- table(trap_extract)[,c("Cotton")]/nrow(trap_extract)
  trap_year$percent_peanuts <- table(trap_extract)[,c("Peanuts")]/nrow(trap_extract)
  trap_year$percent_soybeans <- table(trap_extract)[,c("Soybeans")]/nrow(trap_extract)
  trap_year$percent_sorghum <- table(trap_extract)[,c("Sorghum")]/nrow(trap_extract)
  trap_year$percent_sweetcorn <- table(trap_extract)[,c("Sweet Corn")]/nrow(trap_extract)
  trap_year$percent_tobacco <- table(trap_extract)[,c("Tobacco")]/nrow(trap_extract)
  trap_year$egg_area <- sum(table(trap_extract)[,c("Corn")],
                            table(trap_extract)[,c("Cotton")],
                            table(trap_extract)[,c("Peanuts")],
                            table(trap_extract)[,c("Soybeans")],
                            table(trap_extract)[,c("Sorghum")],
                            table(trap_extract)[,c("Sweet Corn")],
                            table(trap_extract)[,c("Tobacco")])
  assign(paste0("trap_", year), trap_year)
}

trap_vect <- rbind(trap_2020, trap_2021, trap_2022, trap_2023)
trap_vect <- project(trap_vect, "EPSG:4326")
trap_data <- cbind(as.data.frame(trap_vect), long = crds(trap_vect)[,1],
                   lat = crds(trap_vect)[,2])
write_csv(trap_data, file = paste0(output_dir, "/traps_extracted_data.csv"))
```

#Buffer 800
```{r}
for (year in seq(2020, 2023)) {
  cdl_data <- rast(paste0(cdl_path, year, "_30m_cdls/", year, "_30m_cdls.tif"))
  cdl_data <- crop(cdl_data, empty_vect)
  trap_year <- trap_vect[trap_vect$year == year,]
  trap_buffer <- buffer(trap_year, 800)
  trap_extract <- terra::extract(cdl_data, trap_buffer)
  trap_year$percent_corn <- table(trap_extract)[,c("Corn")]/nrow(trap_extract)
  trap_year$percent_cotton <- table(trap_extract)[,c("Cotton")]/nrow(trap_extract)
  trap_year$percent_peanuts <- table(trap_extract)[,c("Peanuts")]/nrow(trap_extract)
  trap_year$percent_soybeans <- table(trap_extract)[,c("Soybeans")]/nrow(trap_extract)
  trap_year$percent_sorghum <- table(trap_extract)[,c("Sorghum")]/nrow(trap_extract)
  trap_year$percent_sweetcorn <- table(trap_extract)[,c("Sweet Corn")]/nrow(trap_extract)
  trap_year$percent_tobacco <- table(trap_extract)[,c("Tobacco")]/nrow(trap_extract)
  trap_year$egg_area <- sum(table(trap_extract)[,c("Corn")],
                            table(trap_extract)[,c("Cotton")],
                            table(trap_extract)[,c("Peanuts")],
                            table(trap_extract)[,c("Soybeans")],
                            table(trap_extract)[,c("Sorghum")],
                            table(trap_extract)[,c("Sweet Corn")],
                            table(trap_extract)[,c("Tobacco")])
  assign(paste0("trap_", year), trap_year)
}

trap_vect <- rbind(trap_2020, trap_2021, trap_2022, trap_2023)
trap_vect <- project(trap_vect, "EPSG:4326")
trap_data <- cbind(as.data.frame(trap_vect), long = crds(trap_vect)[,1],
                   lat = crds(trap_vect)[,2])
write_csv(trap_data, file = paste0(output_dir, "/traps_extracted_data.csv"))
```

#Buffer 1000
```{r}
for (year in seq(2020, 2023)) {
  cdl_data <- rast(paste0(cdl_path, year, "_30m_cdls/", year, "_30m_cdls.tif"))
  cdl_data <- crop(cdl_data, empty_vect)
  trap_year <- trap_vect[trap_vect$year == year,]
  trap_buffer <- buffer(trap_year, 1000)
  trap_extract <- terra::extract(cdl_data, trap_buffer)
  trap_year$percent_corn <- table(trap_extract)[,c("Corn")]/nrow(trap_extract)
  trap_year$percent_cotton <- table(trap_extract)[,c("Cotton")]/nrow(trap_extract)
  trap_year$percent_peanuts <- table(trap_extract)[,c("Peanuts")]/nrow(trap_extract)
  trap_year$percent_soybeans <- table(trap_extract)[,c("Soybeans")]/nrow(trap_extract)
  trap_year$percent_sorghum <- table(trap_extract)[,c("Sorghum")]/nrow(trap_extract)
  trap_year$percent_sweetcorn <- table(trap_extract)[,c("Sweet Corn")]/nrow(trap_extract)
  trap_year$percent_tobacco <- table(trap_extract)[,c("Tobacco")]/nrow(trap_extract)
  trap_year$egg_area <- sum(table(trap_extract)[,c("Corn")],
                            table(trap_extract)[,c("Cotton")],
                            table(trap_extract)[,c("Peanuts")],
                            table(trap_extract)[,c("Soybeans")],
                            table(trap_extract)[,c("Sorghum")],
                            table(trap_extract)[,c("Sweet Corn")],
                            table(trap_extract)[,c("Tobacco")])
  assign(paste0("trap_", year), trap_year)
}

trap_vect <- rbind(trap_2020, trap_2021, trap_2022, trap_2023)
trap_vect <- project(trap_vect, "EPSG:4326")
trap_data <- cbind(as.data.frame(trap_vect), long = crds(trap_vect)[,1],
                   lat = crds(trap_vect)[,2])
write_csv(trap_data, file = paste0(output_dir, "/traps_extracted_data.csv"))
```

```{r}

buffer_lengths <- c(200, 400, 600, 800, 1000, 1200)
for (year in 2020:2023) {
  for (buffer in buffer_lengths) {
  cdl_data <- rast(paste0(cdl_path, year, "_30m_cdls/", year, "_30m_cdls.tif"))
  cdl_data <- crop(cdl_data, empty_vect)
  trap_year <- trap_vect[trap_vect$year == year,]
  trap_extract <- terra::extract(cdl_data, trap_buffer)
  trap_year$percent_corn <- table(trap_extract)[,c("Corn")]/nrow(trap_extract)
  trap_year$percent_cotton <- table(trap_extract)[,c("Cotton")]/nrow(trap_extract)
  trap_year$percent_peanuts <- table(trap_extract)[,c("Peanuts")]/nrow(trap_extract)
  trap_year$percent_soybeans <- table(trap_extract)[,c("Soybeans")]/nrow(trap_extract)
  trap_year$percent_sorghum <- table(trap_extract)[,c("Sorghum")]/nrow(trap_extract)
  trap_year$percent_sweetcorn <- table(trap_extract)[,c("Sweet Corn")]/nrow(trap_extract)
  trap_year$percent_tobacco <- table(trap_extract)[,c("Tobacco")]/nrow(trap_extract)
  trap_year$egg_area <- sum(table(trap_extract)[,c("Corn")],
                            table(trap_extract)[,c("Cotton")],
                            table(trap_extract)[,c("Peanuts")],
                            table(trap_extract)[,c("Soybeans")],
                            table(trap_extract)[,c("Sorghum")],
                            table(trap_extract)[,c("Sweet Corn")],
                            table(trap_extract)[,c("Tobacco")])
  assign(paste0("trap_", year), trap_year)
}
}
trap_vect <- rbind(trap_2020, trap_2021, trap_2022, trap_2023)
trap_vect <- project(trap_vect, "EPSG:4326")
trap_data <- cbind(as.data.frame(trap_vect), long = crds(trap_vect)[,1],
                   lat = crds(trap_vect)[,2])
write_csv(trap_data, file = paste0(output_dir, "/traps_extracted_data.csv"))
```

